---
title: "Sex biased genes"
output: html_document
---

This code will look at some basic stats from the [GTEx database](https://www.gtexportal.org/home/), and will clean and export the database.

## Preliminaries 

Loading libraries

```{r libraries, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```

Loading databases

```{r databases, message=FALSE}
setwd('..')
path <- getwd()
setwd(paste(path, "/DataBases", sep = ""))
samples <- read.csv("SubjSampleSimple.csv")
counts  <- read.table(gzfile("GTEx_Analysis_2016-01-15_v7_RNASeQCv1.1.8_gene_reads.gct.gz"), skip=2, 
                      header=T, row.names=1)
head(samples)
counts[1:6,1:10]
```

## Clean data

The cleaning will involve removing data with no tissue information, and those coming from leukemia cell lines.
Furthermore, we will remove whole tissue samples that have only one sex represented.

```{r cleandata}
samples <- droplevels(filter(samples, SMTSD != '' & SMTSD != "Cells - Leukemia cell line (CML)"))
#Remove tissue with one sex
for(l in 1:length(levels(samples$SMTSD))){
  dat <- filter(samples, SMTSD == levels(samples$SMTSD)[l])
  if(mean(dat$GENDER) == 1 |  mean(dat$GENDER) == 2){
    print(levels(samples$SMTSD)[l])
    samples <- filter(samples, SMTSD != levels(samples$SMTSD)[l])
  }
}
remove(dat)

samples <- droplevels(samples)
keepcounts    <- colnames(counts) %in% as.character(samples$SAMPID)
keepcounts[1] <- TRUE
counts  <- counts[, keepcounts]
samples <- samples[as.character(samples$SAMPID) %in% colnames(counts), ]
```

## Descriptive stats

After cleaning, in this sample, there are `r nlevels(samples$SUBJID)` subjects, representing a total of `r nlevels(samples$SAMPID)` individual samples

```{r counts}
ggplot(samples, aes(as.factor(GENDER))) + geom_bar(aes(fill = AGE), stat="count", position=position_dodge())
```

We can see the tissues and the number of subjects represented by sex.

```{r plot tissues}
#Samples count
#Order table
samplestoplot <- within(samples, SMTSD <- factor(SMTSD, levels=names(sort(table(SMTSD), 
                                                                    decreasing=TRUE))))

median(summary(samples$SMTSD))
ggplot(samplestoplot, aes(SMTSD)) + 
  geom_bar(aes(fill=as.factor(GENDER)), position = position_stack(reverse = TRUE)) + 
  coord_flip() + geom_hline(yintercept = median(summary(samples$SMTSD))) + 
  theme(legend.position = "top")

samplestoplot <- within(samples, SMTS <- factor(SMTS, levels=names(sort(table(SMTS), 
                                                                    decreasing=TRUE))))

median(summary(samples$SMTS))
ggplot(samplestoplot, aes(SMTS)) + 
  geom_bar(aes(fill=as.factor(GENDER)), position = position_stack(reverse = TRUE)) + 
  coord_flip() + geom_hline(yintercept = median(summary(samples$SMTS))) + 
  theme(legend.position = "top")

```


## Saving database

We will now save the cleaned database for future use
```{r saving}
setwd(paste(path, "/Results/GTEx", sep = ""))
write.csv(samples, file = "SubjSample.csv", quote = FALSE, row.names = FALSE)
write.csv(counts, gzfile("GeneReads.csv.gz"), quote = FALSE)
```
