---
title: "Differentially expressed genes per tissue"
output: html_document
---

In this code we will estimate differential expression for each tissue and then export the final list of genes.
Briefly, we will remove genes with low count for each tissue, and run DESeq2 pipeline to estimate differential expression.

## Preliminaries

Loading libraries

```{r libraries, message=FALSE}
library(DESeq2)
library(edgeR)
```

Loading databases

```{r databases, message=FALSE, warning=FALSE}
setwd('..')
path <- getwd()
setwd(paste(path, "/Results/GTEx", sep = ""))
samples <- read.csv("SubjSample.csv")
counts  <- read.csv("GeneReads.csv", row.names=1)
head(samples)
counts[1:4,1:4]
```

## Estimating DE genes

This is the core of the script.
For each tissue we will run the following pipeline:
- First, we will transform the gene counts into counts per million (cpm) values, and remove those with values lower than the n of the minimum group size
- Second, and following the DESeq2 pipeline, we will compute the differential expression analysis, using Age and Sex as covariates
- Finally, a table with the results will be exported 

```{r divide by tissue}
setwd(paste(path, "/Results/DEGenes", sep = ""))

SamplesByTissue <- split(samples, samples$SMTSD)

for(tis in 1:length(SamplesByTissue)){
  GeneCounts <- counts[,colnames(counts) %in% SamplesByTissue[[tis]]$SAMPID]
  keep       <- rowSums(cpm(GeneCounts)>1) >= min(summary(as.factor(SamplesByTissue[[tis]]$GENDER))) # where is the minimum group size 
  GeneCounts <- GeneCounts[keep,] 
  cond       <- SamplesByTissue[[tis]][,3:4] #Create covariate conditions (Gender and age)
  rownames(cond) <- SamplesByTissue[[tis]][,1] 
  cond$GENDER    <- as.factor(cond$GENDER) 
  #DESeq analysis
  mat <- DESeqDataSetFromMatrix(countData = GeneCounts,  #Creating deseq database with counts and covariates (braincond)
                                colData = cond,
                                design  = ~ GENDER + AGE)
  mat <- estimateSizeFactors( mat ) #The following three lines of code are the proper estimation
  mat <- estimateDispersions( mat ) #This takes a while
  mat <- nbinomWaldTest(mat)
  resmat <- results(mat, pAdjustMethod = "BH", contrast = c("GENDER", "1", "2")) #Export the results, with contrast between males and 
  #females. P adjustment can be changed, this case is BH method
  table(resmat$padj < 0.05) #Number of differentially expressed genes
  ResSig <- subset(resmat, padj < 0.05) #Subset results to those differentially expressed only
  save(ResSig, file = paste("resmatSig", as.character(SamplesByTissue[[1]]$SMTSD[1]), ".RData", sep = "_"))
}

```

